---
title: "Seattle Housing Prices: Assesing the effects of acessibility to public goods"
author: "Ahmed Almotaileq, Abigail Edelmen, Christopher Lacrampe"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  html_document:
    code_folding: hide
    highlight: tango
    theme: yeti
    toc: yes
    toc_float: true
---

### Importing the Libraries

```{r, message = FALSE, warning = FALSE}
rm(list=ls(all=TRUE)) # clears working environment
library('tidyverse')  # bundle of packages useful for datamanipulation and organization
library('stargazer')  # package useful in outputting results of statistical models
library('knitr')      # package useful for formatting markdown files to html
library('lmtest')     # has functions for BP Test
library('sandwich')   # has functions for White TestT
library('geosphere')  # has functions for location manipulaton
```

The data for this project was collected from: 

1. Housing data: [Redfin](https://www.redfin.com/city/16163/WA/Seattle) on February 20th, 2018

2. Public goods' locations filtered from [Seattle.gov's](https://data.seattle.gov/Community/My-Neighborhood-Map/82su-5fxf/data) "My Neighborhood Map" collection.
    + Light rail data:   [Seattle.gov](https://data.seattle.gov/Community/Light-Rail-Map/5f4s-t4jf/data) on February 20th, 2018
    + Public parks data: [Seattle.gov](https://data.seattle.gov/Community/Parks-Map/rbbt-rarz) on February 23, 2018.
    + Public schools data: [Seattle.gov](https://data.seattle.gov/Community/Schools-Map/2tje-83f6) on February 23, 2018.
    + Public hospital data: [Seattle.gov](https://data.seattle.gov/Community/Hospitals/khp7-mz6q) on February 23, 2018.
 
### Importing and Cleaning the Data

```{r}
redfinDat <- read.csv("redfin_2018-02-20-12-58-37_Seattle.csv")
lightRailDat <- read.csv("Light_Rail_Map.csv", stringsAsFactors = FALSE)
publicParkDat <- read.csv("Parks_Map.csv", stringsAsFactors = FALSE)
publicSchoolDat <- read.csv("Schools_Map.csv", stringsAsFactors = FALSE)
publicHospitalDat <- read.csv("Hospitals.csv", stringsAsFactors = FALSE)

# Removing uneeded colums
redfinDat <- subset(redfinDat, select = -c(SOLD.DATE,NEXT.OPEN.HOUSE.START.TIME,NEXT.OPEN.HOUSE.END.TIME, URL..SEE.http...www.redfin.com.buy.a.home.comparative.market.analysis.FOR.INFO.ON.PRICING., SOURCE, MLS.,FAVORITE, INTERESTED))

# Adds UW light rail station
uwStation <- list(FEATUREITEM_ID = NA, City.Feature = "Light Rail", Common.Name = as.factor("UW Station"), Address = NA, Website = NA, Longitude = -122.3038, Latitude = 47.6498, Location = "(47.6498, 122.3038)")
lightRailDat = rbind(lightRailDat, uwStation, stringsAsFactors=FALSE)
```

##### Special thanks to stack overflow user eclark for providing the framework for the functions used to calculate minimum distances. See his response to a simmilair query from  [this stackoverflow thread.](https://stackoverflow.com/questions/31732281/finding-minimum-distance-between-two-sets-of-points-in-two-sets-of-r)
#### Calculate the distance from the property location to the nearest light rail station using the following code:

```{r, warning = FALSE, results = "hide"}
# Preparing lat/lon dataframes for function
redfinCoords <- data.frame(location_id=redfinDat$ADDRESS,LATITUDE=redfinDat$LATITUDE,LONGITUDE=redfinDat$LONGITUDE)
lightRailCoords <- data.frame(location_id=lightRailDat$Common.Name,LATITUDE=lightRailDat$Latitude,LONGITUDE=lightRailDat$Longitude)

# Setting up DistLinkFun to find distance from property to closest lightrail station

DistLinkFun <- function(ID){
 TMP <- redfinCoords[redfinCoords$location_id==ID,]
 TMP1 <- distGeo(TMP[,3:2],lightRailCoords[,3:2]) # uses distGeo() function from geosphere package to calculate dist from lat and lon
 TMP2 <- data.frame(redfinCoordsID=ID,lightRailCoordsID=lightRailCoords[which.min(TMP1),1],distanceToLink=min(TMP1)) 
 print(ID)
 return(TMP2)
}

# Distance output of DistLinkFun parameters as redfinCoords$location_id, output is in meters
DistanceLinkMatrix <- rbind_all(lapply(redfinCoords$location_id, DistLinkFun))

# Taking distance variable and adding to original redfinDat dataframe
redfinDat$distanceToLink <- DistanceLinkMatrix$distanceToLink # values are in meters

# Converting distance to miles
redfinDat$distanceToLink <- redfinDat$distanceToLink*0.000621371 # num is rate of meters per mile

# Adding the closest lightRail station variable to redfinDat
redfinDat$ClosestStation <- DistanceLinkMatrix$lightRailCoordsID
```

#### Calculate the distance from the property location to the nearest public park using the following code:

```{r, warning = FALSE, results = "hide"}
# preparing lat/lon dataframe for function
publicParkCoords <- data.frame(location_id=publicParkDat$Common.Name,LATITUDE=publicParkDat$Latitude,LONGITUDE=publicParkDat$Longitude)

# Setting up DistParkFun to find distance from property to closest public park

DistParkFun <- function(ID){
 TMP <- redfinCoords[redfinCoords$location_id==ID,]
 TMP1 <- distGeo(TMP[,3:2],publicParkCoords[,3:2]) # uses distGeo() function from geosphere package to calculate dist from lat and lon
 TMP2 <- data.frame(redfinCoordsID=ID,publicParkCoordsID=publicParkCoords[which.min(TMP1),1],distanceToPark=min(TMP1)) 
 print(ID)
 return(TMP2)
}

# Distance output of DistFun parameters as redfinCoords$location_id, output is in meters
DistanceParkMatrix <- rbind_all(lapply(redfinCoords$location_id, DistParkFun))

# Taking distance variable and adding to original redfinDat dataframe
redfinDat$distanceToPark <- DistanceParkMatrix$distanceToPark # values are in meters

# Converting distance to miles
redfinDat$distanceToPark <- redfinDat$distanceToPark*0.000621371 # num is rate of meters per mile

# Adding the public park name variable to redfinDat
redfinDat$ClosestPark <- DistanceParkMatrix$publicParkCoordsID
```

#### Calculate the distance from the property location to the nearest public school using the following code:

```{r, warning = FALSE, results = "hide"}
# preparing lat/lon dataframe for function
publicSchoolCoords <- data.frame(location_id=publicSchoolDat$Common.Name,LATITUDE=publicSchoolDat$Latitude,LONGITUDE=publicSchoolDat$Longitude)

# Setting up DistSchoolFun to find distance from property to closest public school

DistSchoolFun <- function(ID){
 TMP <- redfinCoords[redfinCoords$location_id==ID,]
 TMP1 <- distGeo(TMP[,3:2],publicSchoolCoords[,3:2]) # uses distGeo() function from geosphere package to calculate dist from lat and lon
 TMP2 <- data.frame(redfinCoordsID=ID,publicSchoolCoordsID=publicSchoolCoords[which.min(TMP1),1],distanceToSchool=min(TMP1)) 
 print(ID)
 return(TMP2)
}

# Distance output of DistFun parameters as redfinCoords$location_id, output is in meters
DistanceSchoolMatrix <- rbind_all(lapply(redfinCoords$location_id, DistSchoolFun))

# Taking distance variable and adding to original redfinDat dataframe
redfinDat$distanceToSchool <- DistanceSchoolMatrix$distanceToSchool # values are in meters

# Converting distance to miles
redfinDat$distanceToSchool <- redfinDat$distanceToSchool*0.000621371 # num is rate of meters per mile

# Adding the school name variable to redfinDat
redfinDat$ClosestSchool <- DistanceSchoolMatrix$publicSchoolCoordsID
```

#### Calculate the distance from the property location to the nearest public hospital using the following code:

```{r, warning = FALSE, results = "hide"}
# preparing lat/lon dataframe for function
publicHospitalCoords <- data.frame(location_id=publicHospitalDat$Common.Name,LATITUDE=publicHospitalDat$Latitude,LONGITUDE=publicHospitalDat$Longitude)

# Setting up DistHopsitalFun to find distance from property to closest public hospital

DistHospitalFun <- function(ID){
 TMP <- redfinCoords[redfinCoords$location_id==ID,]
 TMP1 <- distGeo(TMP[,3:2],publicHospitalCoords[,3:2]) # uses distGeo() function from geosphere package to calculate dist from lat and lon
 TMP2 <- data.frame(redfinCoordsID=ID,publicHospitalCoordsID=publicHospitalCoords[which.min(TMP1),1],distanceToHospital=min(TMP1)) 
 print(ID)
 return(TMP2)
}

# Distance output of DistFun parameters as redfinCoords$location_id, output is in meters
DistanceHospitalMatrix <- rbind_all(lapply(redfinCoords$location_id, DistHospitalFun))

# Taking distance variable and adding to original redfinDat dataframe
redfinDat$distanceToHospital <- DistanceHospitalMatrix$distanceToHospital # values are in meters

# Converting distance to miles
redfinDat$distanceToHospital <- redfinDat$distanceToHospital*0.000621371 # num is rate of meters per mile

# Adding the hospital name variable to redfinDat
redfinDat$ClosestHospital <- DistanceHospitalMatrix$publicHospitalCoordsID
```