---
title: "Seattle Housing Prices and Access to the Light Rail"
author: "Abigail Edelmen, Ahmed Almotaileq, Christopher Lacrampe"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  html_document:
    code_folding: hide
    highlight: tango
    theme: yeti
    toc: yes
---

#### Importing the Libraries

```{r, message = FALSE, warning = FALSE}
rm(list=ls(all=TRUE)) # clears working environment
library('tidyverse')  # bundle of packages useful for datamanipulation and organization
library('stargazer')  # package useful in outputting results of statistical models
library('knitr')      # package useful for formatting markdown files to html
library('lmtest')     # has functions for BP Test
library('sandwich')   # has functions for White TestT
library('geosphere')  # has functions for location manipulaton
```

The data for this project was collected from [Redfin](https://www.redfin.com/city/16163/WA/Seattle) and [Seattle.gov](https://data.seattle.gov/Community/Light-Rail-Map/5f4s-t4jf/data) on February 20th, 2018. These series include...
 
#### Importing and Cleaning the Dataset

```{r}
redfinDat <- read.csv("redfin_2018-02-20-12-58-37_Seattle.csv")
lightRailDat <- read.csv("Light_Rail_Map.csv", stringsAsFactors = FALSE)

# Removing uneeded colums
redfinDat <- subset(redfinDat, select = -c(SOLD.DATE,NEXT.OPEN.HOUSE.START.TIME,NEXT.OPEN.HOUSE.END.TIME, URL..SEE.http...www.redfin.com.buy.a.home.comparative.market.analysis.FOR.INFO.ON.PRICING., SOURCE, MLS.,FAVORITE, INTERESTED))

# Adds UW light rail station
uwStation <- list(FEATUREITEM_ID = NA, City.Feature = "Light Rail", Common.Name = as.factor("UW Station"), Address = NA, Website = NA, Longitude = -122.3038, Latitude = 47.6498, Location = "(47.6498, 122.3038)")
lightRailDat = rbind(lightRailDat, uwStation, stringsAsFactors=FALSE)
```

##### Calculate the distance from the property location to the nearest light rail station using the following code:
###### Special thanks to stack overflow user eclark. See his response to a simmilair query from  [this](https://stackoverflow.com/questions/31732281/finding-minimum-distance-between-two-sets-of-points-in-two-sets-of-r) thread.

```{r, warning = FALSE, results = "hide"}
# Preparing lat/lon databases for function
redfinCoords <- data.frame(location_id=redfinDat$ADDRESS,LATITUDE=redfinDat$LATITUDE,LONGITUDE=redfinDat$LONGITUDE)
lightRailCoords <- data.frame(location_id=lightRailDat$Common.Name,LATITUDE=lightRailDat$Latitude,LONGITUDE=lightRailDat$Longitude)

# Setting up DistFun to find distance from property to closes lightrail station

DistFun <- function(ID){
 TMP <- redfinCoords[redfinCoords$location_id==ID,]
 TMP1 <- distGeo(TMP[,3:2],lightRailCoords[,3:2]) # uses distGeo() function from geosphere package to calculate dist from lat and lon
 TMP2 <- data.frame(redfinCoordsID=ID,lightRailCoordsID=lightRailCoords[which.min(TMP1),1],DistanceBetween=min(TMP1)) 
 print(ID)
 return(TMP2)
}

# Distance output of DistFun parameters as redfinCoords$location_id, output is in meters
DistanceMatrix <- rbind_all(lapply(redfinCoords$location_id, DistFun))

# Taking distance variable and adding to original redfinDat dataframe
redfinDat$distance <- DistanceMatrix$DistanceBetween # values are in meters

# Converting distance to miles
redfinDat$distance <- redfinDat$distance*0.000621371 # num is rate of meters per mile

# Adding the lightRail station variable
redfinDat$station <- DistanceMatrix$lightRailCoordsID
```


